---
title: "ABCD Demographics"
author: "Eric W. Bridgeford"
date: "2024-07-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
require(tidyverse)
require(dplyr)
require(ggplot2)
require(jsonlite)
require(ggridges)
require(transport)
require(patchwork)
```

# ABCD Demographics plot (cross-site)

Raw data taken from "DCAN Labs ABCD-BIDS Community Collection (ABCC)"; file "participants_v1.0.3."

```{r}
demo_key <- read_json("../data/abcd/demo_key.json")

site_key <- unlist(demo_key$site$Levels)
match_key <- unlist(demo_key$matched_group$Levels)
sex_key <- unlist(demo_key$sex$Levels)

raw_demo_dat <- read_tsv("../data/abcd/demographics.tsv")
```

```{r}
# pre-process the data using the demographics key
demo_dat <- raw_demo_dat %>%
  rename("Latinx" = `Do you consider the child Hispanic/Latino/Latina?`,
         "Black"=`Black/African American`) %>%
  mutate(
    White = ifelse(White == 1, 1, ifelse(White == 0, 0, NA)),
    Black = ifelse(Black == 1, 1, ifelse(Black == 0, 0, NA)),
    Chinese = ifelse(Chinese == 1, 1, ifelse(Chinese == 0, 0, NA)),
    Vietnamese = ifelse(Vietnamese == 1, 1, ifelse(Vietnamese == 0, 0, NA)),
    Filipino = ifelse(Filipino == 1, 1, ifelse(Filipino == 0, 0, NA)),
    Japanese = ifelse(Japanese == 1, 1, ifelse(Japanese == 0, 0, NA)),
    `Other Asian` = ifelse(`Other Asian` == 1, 1, ifelse(`Other Asian` == 0, 0, NA)),
    pc1=ifelse(pc1 != 888, pc1, NA),
    pc2=ifelse(pc2 != 888, pc2, NA),
    pc3=ifelse(pc3 != 888, pc3, NA),
    parental_education=ifelse(parental_education %in% c(999, 777, 888), NA, parental_education),
    Korean = ifelse(Korean == 1, 1, ifelse(Korean == 0, 0, NA)),
    income = ifelse(income %in% c(777, 888, 999), NA, income),
    anesthesia_exposure=ifelse(anesthesia_exposure == 888, NA, ifelse(anesthesia_exposure > 0, 1, 0)),
    site = fct_recode(factor(site), !!!setNames(names(site_key), site_key)),
    male = ifelse(sex == 1, 1, ifelse(sex %in% c(2, 3, 4), 0, NA)),
    matched_group = factor(matched_group),
    Latinx=ifelse(Latinx == 1, 1, ifelse(Latinx == 2, 0, NA)),
    handedness=ifelse(handedness == 1, 1, ifelse(handedness %in% c(2, 3), 0, NA))
  ) %>%
  filter(grepl("ses-baseline", session_id), matched_group %in% c(1, 2, 3), site != "Not reported", site != "Unknown") %>%
  mutate(Asian = as.numeric(Chinese + Vietnamese + Filipino + Japanese + `Other Asian` + Korean > 0)) %>%
  rename("Right-handed"=handedness) %>%
  select(participant_id, site, male, White, Black, Asian, Latinx, age, income, anesthesia_exposure, pc1, pc2, pc3, `Right-handed`, parental_education)
```

```{r}
normalized_data <- demo_dat %>%
  pivot_longer(cols = male:parental_education,
               names_to = "variable",
               values_to = "value") %>%
  group_by(variable) %>%
  mutate(upper.q = quantile(value, probs=0.95, na.rm=TRUE), lower.q = quantile(value, probs=0.05, na.rm=TRUE)) %>%
  # compute the mean per-variable by site, excluding NAs
  group_by(site, variable, upper.q, lower.q) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  # normalize the means per-site
  mutate(normalized_mean = (mean_value - lower.q) / (upper.q - lower.q)) %>%
  select(site, variable, normalized_mean) %>%
  mutate(variable = as.character(variable)) %>%
  mutate(variable = factor(variable, 
                           levels = c("male", "White", "Black", "Asian", "Latinx",
                                      "age", "pc1", "pc2", "pc3", "parental_education",
                                      "income", "anesthesia_exposure", "Right-handed"), 
                           ordered = TRUE)) %>%
  mutate(variable = recode_factor(variable,  
                                  "male" = "Percent biological male",
                                  "White"="Percent White",
                                  "Black"="Percent Black",
                                  "Asian"="Percent Asian",
                                  "Latinx"="Percent Latinx",
                                  "age" = "Age",
                                  "pc1" = "General ability PC", 
                                  "pc2" = "Exec. function PC", 
                                  "pc3" = "Learning/memory PC",
                                  "parental_education" = "Parental education", 
                                  "income" = "Parental income",
                                  "anesthesia_exposure" = "Any anesthesia exposure"))

dset_avg_norm <- demo_dat %>%
  pivot_longer(cols = male:parental_education,
               names_to = "variable",
               values_to = "value") %>%
  group_by(variable) %>%
  mutate(upper.q = quantile(value, probs=0.95, na.rm=TRUE), lower.q = quantile(value, probs=0.05, na.rm=TRUE)) %>%
  group_by(variable, upper.q, lower.q) %>%
  summarize(mean_value=mean(value, na.rm=TRUE), .groups="drop") %>%
  # normalize the means
  mutate(normalized_mean = (mean_value - lower.q) / (upper.q - lower.q)) %>%
  mutate(variable = factor(variable, 
                           levels = c("male", "White", "Black", "Asian", "Latinx",
                                      "age", "pc1", "pc2", "pc3", "parental_education",
                                      "income", "anesthesia_exposure", "Right-handed"), 
                           ordered = TRUE)) %>%
  mutate(variable = recode_factor(variable,  
                                  "male" = "Percent biological male",
                                  "White"="Percent White",
                                  "Black"="Percent Black",
                                  "Asian"="Percent Asian",
                                  "Latinx"="Percent Latinx",
                                  "age" = "Age",
                                  "pc1" = "General ability PC", 
                                  "pc2" = "Exec. function PC", 
                                  "pc3" = "Learning/memory PC",
                                  "parental_education" = "Parental education", 
                                  "income" = "Parental income",
                                  "anesthesia_exposure" = "Any anesthesia exposure"))
```

```{r, fig.height=5, fig.width=9}
normalized_data %>%
  ggplot(aes(x = normalized_mean, y = variable)) +
    geom_point(aes(color = site, group = site), alpha = 0.5, position =  position_dodge(width = 0.2)) +
    geom_point(data=dset_avg_norm, color="black", size=8, shape="+") +
    scale_x_continuous(limits = c(0, 1), expand = c(0, 0), name = "Normalized mean (per-site)") +
    scale_y_discrete(name = "Demographic covariate") +
    #ggtitle("Differences in ABCD demographics (across sites)") +
    theme_bw() +
    theme(legend.position = "right", panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
          text=element_text(size=15))
```

## Statistical Analyses

```{r}
# for 2xK contingency tables, use Fisher-Freeman-Halton
ffh.test <- function(predictors, groups) {
  cont_tbl <- table(predictors, groups)
  # 10k null replicates
  res <- fisher.test(cont_tbl, simulate.p.value = TRUE, B = 10000)
  return(list(Statistic=res$estimate, p.value=res$p.value, Test="FFH Test"))
}

# for continuous data, use KW
kw.test <- function(predictors, groups) {
  res <- kruskal.test(predictors ~ groups)
  return(list(Statistic=res$statistic, p.value=res$p.value, Test="KW Test"))
}

excluded.tests <- list("Right-handed"=ffh.test, "anesthesia_exposure"=ffh.test,
                       "income"=kw.test, "parental_education"=kw.test, "pc3"=kw.test,
                       "pc2"=kw.test, "pc1"=kw.test, "age"=kw.test, "Latinx"=ffh.test, "Asian"=ffh.test,
                       "Black"=ffh.test, "White"=ffh.test, "male"=ffh.test)

test.outcomes.site <- do.call(rbind, lapply(names(excluded.tests), function(covar.name) {
  out <- do.call(excluded.tests[[covar.name]], list(demo_dat[[covar.name]], demo_dat$site))
  out$Covariate <- covar.name
  out
})) %>%
  data.frame() %>%
  select(Covariate, Test, Statistic, p.value)
```

# ABCD ADHD Exploration

Raw data taken from [here](https://nda.nih.gov/general-query.html?q=query=featured-datasets:Adolescent%20Brain%20Cognitive%20Development%20Study%20(ABCD)), by obtaining "Release 5.0 Raw Data - June 2023 (4 GB)".

```{r}
cbcl_raw <- read.csv("../data/abcd/mh_cbcl.csv")
motion_raw <- read.csv("../data/abcd/mri_motion.csv")
```

## CBCL survey

Key taken from [here](https://data-dict.abcdstudy.org/?).

Key:
+ cbcl_scr_dsm5_adhd_r: Recommended ADHD CBCL DSM5 Scale (raw score)
+ cbcl_scr_dsm5_adhd_t: Recommended ADHD CBCL DSM5 Scale (t-score)
+ cbcl_scr_dsm5_adhd_m: Recommended ADHD CBCL DSM5 Scale (missing values)		
+ cbcl_scr_dsm5_adhd_nm: Recommended ADHD CBCL DSM5 Scale (number of missing values)
+ rsfmri_meanmotion: Resting state fMRI - Average framewise displacement in mm
+ rsfmri_subthreshnvols: Resting state fMRI - Number of frames with FD < 0.2
+ rsfmri_subthreshcontignvols: Resting state fMRI - Number of frames with FD < 0.2 and at least 5 contiguous, supra-threshold frames

```{r}
cbcl_dat <- cbcl_raw %>%
  filter(eventname == "baseline_year_1_arm_1") %>%
  select(src_subject_id, cbcl_scr_dsm5_adhd_t, cbcl_scr_dsm5_adhd_r, cbcl_scr_dsm5_adhd_nm, cbcl_scr_dsm5_adhd_m) %>%
  rename(participant_id=src_subject_id, ADHD.raw=cbcl_scr_dsm5_adhd_r, ADHD.tscore=cbcl_scr_dsm5_adhd_t, ADHD.mv=cbcl_scr_dsm5_adhd_m,
         ADHD.nmissing=cbcl_scr_dsm5_adhd_nm) %>%
  mutate(participant_id = str_replace(participant_id, "^([A-Z]+)_([A-Z]+.*)$", "sub-\\1\\2"))

motion_dat <- motion_raw %>%
  filter(eventname == "baseline_year_1_arm_1") %>%
  select(src_subject_id, rsfmri_meanmotion, rsfmri_subthreshnvols, rsfmri_subthreshcontignvols) %>%
  rename(participant_id=src_subject_id, MeanMotion=rsfmri_meanmotion, NVolumes=rsfmri_subthreshnvols,
         NVolsContiguous=rsfmri_subthreshcontignvols) %>%
  mutate(participant_id = str_replace(participant_id, "^([A-Z]+)_([A-Z]+.*)$", "sub-\\1\\2"))

# merge in the adhd annotation data with the original demographic data
merge_demo_cbcl_motion <- demo_dat %>%
  left_join(cbcl_dat, by=c("participant_id"="participant_id")) %>%
  left_join(motion_dat, by=c("participant_id"="participant_id"))
```

## Visualizations

```{r}
cbcl.raw.plt <- merge_demo_cbcl_motion %>%
  ggplot(aes(ADHD.raw)) +
    geom_histogram(binwidth=1, color="black", fill="white") +
    scale_y_continuous(name="Number of Individuals") +
    scale_x_continuous(name="Recommended ADHD CBCL DSM5 Scale (raw score)") +
    ggtitle("(A) Child Behavior Checklist Raw Scores") +
    theme_bw() +
    theme(legend.position = "right", panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
          text=element_text(size=15))
```

```{r}
sum.merged <- merge_demo_cbcl_motion %>%
  group_by(ADHD.raw) %>%
  summarize(N.MeanMotion =sum(!is.na(MeanMotion)), SE.MeanMotion=sd(MeanMotion, na.rm=TRUE)/sqrt(N.MeanMotion), Mean.MeanMotion=mean(MeanMotion, na.rm=TRUE),
            N.NVolumes=sum(!is.na(NVolumes)), SE.NVolumes=sd(NVolumes, na.rm=TRUE)/sqrt(N.NVolumes), Mean.NVolumes=mean(NVolumes, na.rm=TRUE),
            N.NVolsCont =sum(!is.na(NVolsContiguous)), SE.NVolsCont=sd(NVolsContiguous, na.rm=TRUE)/sqrt(N.NVolsCont), Mean.NVolsCont=mean(NVolsContiguous, na.rm=TRUE)) %>%
  ungroup() %>%
  pivot_longer(
    cols = -ADHD.raw,
    names_to = c(".value", "Measure"),
    names_pattern = "(N|SE|Mean)\\.(.+)"
  )
```

```{r}
plt.excl_criteria <- sum.merged %>%
  mutate(Measure=recode_factor(Measure, "MeanMotion"="I. Mean FD (mm)", "NVolumes"="II. Volumes < 0.2 mm FD")) %>%
  filter(Measure != "NVolsCont") %>%
  ggplot(aes(x=ADHD.raw)) +
    geom_point(aes(y=Mean)) +
    geom_errorbar(aes(ymin = Mean - SE, ymax=Mean + SE)) +
    scale_x_continuous(name="Recommended ADHD CBCL DSM5 Scale (raw score)") +
    scale_y_continuous(name="Average framewise displacement (mm)") +
    facet_grid(Measure~., scale="free_y", switch = "y") +
    theme_bw() +
    ggtitle("(B) Selection differences by ADHD diagnostic score") +
    theme(legend.position = "right", panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
          strip.text.y.left = element_text(angle = 90, hjust = 0.5, size=15), axis.title.y = element_blank(),
          text=element_text(size=15), strip.background = element_blank(), strip.placement = "outside")
```

```{r}
# exclusion criterion of 0.5 mm
excl_mean = 0.5

merged_covar_excluded_normdat <- (merged_covar_excluded <- merge_demo_cbcl_motion %>%
  mutate(Sample=ifelse(MeanMotion < 0.5, "Incl. (Mean FD < 0.5)", ifelse(!is.na(MeanMotion), "Excl. (Mean FD > 0.5)", NA)))) %>%
  filter(!is.na(Sample)) %>%
  select(-c("MeanMotion", "NVolumes", "NVolsContiguous", "ADHD.tscore", "ADHD.nmissing", "ADHD.mv")) %>%
  pivot_longer(cols = male:ADHD.raw,
               names_to = "variable",
               values_to = "value") %>%
  group_by(variable) %>%
  mutate(upper.q = quantile(value, probs=0.95, na.rm=TRUE), lower.q = quantile(value, probs=0.05, na.rm=TRUE)) %>%
  # compute the mean per-variable by site, excluding NAs
  group_by(Sample, variable, upper.q, lower.q) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  # normalize the means per-site
  mutate(normalized_mean = (mean_value - lower.q) / (upper.q - lower.q)) %>%
  select(Sample, variable, normalized_mean) %>%
  mutate(variable = as.character(variable)) %>%
  mutate(variable = factor(variable, 
                           levels = c("male", "White", "Black", "Asian", "Latinx",
                                      "age", "pc1", "pc2", "pc3", "parental_education",
                                      "income", "anesthesia_exposure", "Right-handed",
                                      "ADHD.raw"), 
                           ordered = TRUE)) %>%
  mutate(variable = recode_factor(variable,  
                                  "male" = "Percent biological male",
                                  "White"="Percent White",
                                  "Black"="Percent Black",
                                  "Asian"="Percent Asian",
                                  "Latinx"="Percent Latinx",
                                  "age" = "Age",
                                  "pc1" = "General ability PC", 
                                  "pc2" = "Exec. function PC", 
                                  "pc3" = "Learning/memory PC",
                                  "parental_education" = "Parental education", 
                                  "income" = "Parental income",
                                  "anesthesia_exposure" = "Any anesthesia exposure",
                                  "Right-handed"="Right-handed",
                                  "ADHD.raw"="ADHD CBCL DSM5 Scale"))

dset_avg_norm_adhd <- merge_demo_cbcl_motion %>%
  select(-c("MeanMotion", "NVolumes", "NVolsContiguous", "ADHD.tscore", "ADHD.nmissing", "ADHD.mv")) %>%
  pivot_longer(cols = male:ADHD.raw,
               names_to = "variable",
               values_to = "value") %>%
  group_by(variable) %>%
  mutate(upper.q = quantile(value, probs=0.95, na.rm=TRUE), lower.q = quantile(value, probs=0.05, na.rm=TRUE)) %>%
  group_by(variable, upper.q, lower.q) %>%
  summarize(mean_value=mean(value, na.rm=TRUE), .groups="drop") %>%
  # normalize the means
  mutate(normalized_mean = (mean_value - lower.q) / (upper.q - lower.q)) %>%
  mutate(variable = factor(variable, 
                           levels = c("male", "White", "Black", "Asian", "Latinx",
                                      "age", "pc1", "pc2", "pc3", "parental_education",
                                      "income", "anesthesia_exposure", "Right-handed",
                                      "ADHD.raw"), 
                           ordered = TRUE)) %>%
  mutate(variable = recode_factor(variable,  
                                  "male" = "Percent biological male",
                                  "White"="Percent White",
                                  "Black"="Percent Black",
                                  "Asian"="Percent Asian",
                                  "Latinx"="Percent Latinx",
                                  "age" = "Age",
                                  "pc1" = "General ability PC", 
                                  "pc2" = "Exec. function PC", 
                                  "pc3" = "Learning/memory PC",
                                  "parental_education" = "Parental education", 
                                  "income" = "Parental income",
                                  "anesthesia_exposure" = "Any anesthesia exposure",
                                  "Right-handed"="Right-handed",
                                  "ADHD.raw"="ADHD CBCL DSM5 Scale"))
```

```{r}
covar.tbl.incl_excl <- merged_covar_excluded_normdat %>%
  ggplot(aes(x = normalized_mean, y = variable)) +
    geom_point(aes(color = Sample, group = Sample), alpha = 0.7, size=3) +
    geom_point(data=dset_avg_norm_adhd, color="black", size=8, shape="+") +
    scale_x_continuous(limits = c(0, 1), expand = c(0, 0), name = "Normalized mean (by incl./excl. criteria)") +
    scale_y_discrete(name = "Demographic covariate") +
    ggtitle("(C) Differences in Demographics (by Mean FD group)") +
    theme_bw() +
    theme(legend.position = "right", panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
          text=element_text(size=15))
```

```{r, fig.height=5, fig.width=21}
grid.arrange(cbcl.raw.plt, plt.excl_criteria, covar.tbl.incl_excl, nrow=1, widths=c(0.7, 0.7, 1))
```

## Statistical Analyses

```{r}
# for 2x2 contingency tables, use Fisher exact
fe.test <- function(predictors, groups) {
  cont_tbl <- table(predictors, groups)
  res <- fisher.test(cont_tbl, alternative="two.sided")
  return(list(Statistic=res$estimate, p.value=res$p.value, Test="Fisher Exact"))
}

# for continuous data, use Mann-Whitney U-Test
u.test <- function(predictors, groups) {
  res <- wilcox.test(predictors ~ groups, alternative="two.sided")
  return(list(Statistic=res$statistic, p.value=res$p.value, Test="U-Test"))
}

excluded.tests <- list("ADHD.raw"=u.test, "Right-handed"=fe.test, "anesthesia_exposure"=fe.test,
                       "income"=u.test, "parental_education"=u.test, "pc3"=u.test,
                       "pc2"=u.test, "pc1"=u.test, "age"=u.test, "Latinx"=fe.test, "Asian"=fe.test,
                       "Black"=fe.test, "White"=fe.test, "male"=fe.test)

test.outcomes.adhd_group <- do.call(rbind, lapply(names(excluded.tests), function(covar.name) {
  out <- do.call(excluded.tests[[covar.name]], list(merged_covar_excluded[[covar.name]], merged_covar_excluded$Sample))
  out$Covariate <- covar.name
  out
})) %>%
  data.frame() %>%
  select(Covariate, Test, Statistic, p.value)
```

# Multiple comparisons corrections

```{r}
test.outcomes <- rbind(test.outcomes.site %>% mutate(Group="Site"),
                       test.outcomes.adhd_group %>% mutate(Group="Motion")) %>%
  ungroup() %>%
  mutate(p.value=p.adjust(p.value, method="BH"))

kable(test.outcomes)
```

# Counts of each group

```{r}
print(table(demo_dat$site))

print(table(merged_covar_excluded$Sample))
```

